---
title: "R scripts"
author: marcello gallucci
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

In this page one can found all the R functions discussed in Perugini, Gallucci, Constantini (2018), A practical primer to power analysis for simple experimental designs. __International Review of Social Psychology__.  [rewf here](here). Examples are taken from the papers. In all the examples that requires multiple lines of code the user needs to update only the variables listed before `### end of input ###` comment. The results are computed automatically for the required N. Small changes in the code are required for other power parameters.

# t-test
## Independent Samples

The simplest function for independent samples t.test. To use with Cohen's d, set `sd=1` and put `delta=d`, where `d` is your effect size.

```{r }

power.t.test(power=.80, sig.level=.05, delta=.5, sd=1)

```
Required n for each group should be rounded up to the first whole number. Exactly one of `n`, `delta`, `sd`, `power`, and `sig.level` must be NULL. The NULL parameter is estimated by the function. Thus, post-hoc analysis is done omitting `power` and including `n`. 

```{r}

power.t.test(n=64, sig.level=.05, delta=.5, sd=1)

```
`?power.t.test` for all the other options.


## Sensitivity analysis

In the paper we suggested to look at the relationship between total sample size and effect size for a given power level (.80 in the example) and alpha (.05 in the example). In R this can be done as follows:

Same function with the option `type = "paired"`
```{r }
library(ggplot2)

N<-seq(20,140,by=20)

## end of input ##
dat<-data.frame(n=N)
for (i in seq(N)) {
   one.power<-power.t.test(power=.80, sig.level=.05, sd=1, n=N[i])
   dat$d[i]<-one.power$delta
}
dat
plt<-ggplot(dat,aes(y=d,x=n, label = round(d,digits = 2))) +stat_smooth(method = "loess",se=FALSE)
plt<-plt+labs(x="Totale sample size",y="Effect size d")
plt<-plt+scale_x_continuous(breaks=seq(min(N),max(N),by=20))
plt<-plt+geom_label()
plt


```



## Paired Samples

Same function with the option `type = "paired"`
```{r }

power.t.test(power=.80, sig.level=.05, delta=0.527, sd=1,type = "paired")

```

# One way Analysis of Variance

For all F-test related power analysis, it is generally better to use `pwr.f2.test()` function from [pwr packae](https://cran.r-project.org/web/packages/pwr/index.html). The function uses $f^2$ as effect size, which is the squared of the $f$ used by GPower "ANOVA: Fixed effects, omnibus and one-way". Here we assume one has the $\eta_p^2$, which in one-way ANOVA is the $R^2$. For required N (total sample size) one needs to add $k$ to the 

```{r }
library(pwr)

## input ##
etap<-.326
k<-8 # number of groups in the design
## end of input ##
f2=etap/(1-etap)
dfn<-k-1 # numerator degrees of freedom

(res<-pwr.f2.test(u = dfn,f2 = f2,sig.level = .05,power = .80))
## required N ##
ceiling(res$v)+res$u+1

```

# Factorial Designs

"Assume that in a 3 X 2 factorial design the researcher expects the interaction to explain around 10% of the variance. If the researcher expects no main effect, the proportion of residual variance is 1-.10=.90, so the pη2=.10." 

## method 1

Here we keep using `pwr.f2.test` because it can be used for any F-test in the linear model. However, for prospective power (required total N) the function returns the required error degrees of freedom, so the required N must be approximate adding the effects degrees of freedom to the error degrees of freedom. The following code does it automatically.

```{r }
library(pwr)

## input for interaction##
etap<-.10
k1<-3 # levels of the first factor
k2<-2 # levels of the second factor
## end of input ##

(f2=etap/(1-etap))
dfn<-(k1-1)*(k2-1) # df for the interaction

(res<-pwr.f2.test(u = dfn,f2 = f2,sig.level = .05,power = .80))

## required N ##
ceiling(res$v)+(k1*k2)-1


## input for main effects##
etap<-.10
k1<-3 # levels of the first factor
k2<-2 # levels of the second factor
## end of input ##

(f2=etap/(1-etap))
dfn<-(k1-1) # df for the main effect 1
(res<-pwr.f2.test(u = dfn,f2 = f2,sig.level = .05,power = .80))

## required N ##
ceiling(res$v)+(k1*k2)-1

dfn<-(k2-1) # df for the main effect 2
(res<-pwr.f2.test(u = dfn,f2 = f2,sig.level = .05,power = .80))

## required N ##
ceiling(res$v)+(k1*k2)-1


```

## Method 2

The package [easypower](https://cran.r-project.org/web/packages/easypower/index.html) provides short cuts for factorial ANOVA power analysis. It provides a dedicated function `n.multiway()` which simplifies computation of required N for main effects and interaction. It does not provide estimates for post-hoc power and other applications of power analysis.

```{r }

library(easypower)
## input ##
etap<-.10
k1<-3
k2<-2
## end of input ##

main.eff1 <- list(name = "A", levels = k1, eta.sq = etap)
main.eff2 <- list(name = "C", levels = k2, eta.sq = etap)
int.eff1 <- list(name = "A*C", eta.sq = etap)
n.multiway(iv1 = main.eff1, iv2 = main.eff2, int1 = int.eff1)


```
Results are the same than _method 1_ a part from rounding.

## Power analysis for contrasts

### Generic Contrasts

Main effects and interaction for example in Table 2. They are all equivalent, so we go for the interaction. We compute the $f$ as in the paper, but use $f^2$ as required by `pwr.f2.

```{r }

## input ##
means<-c(10,0,0,0)
sd=5
contInt<-c(1,-1,-1,1)

## end input ##

### contrast value
cvInt<-contInt%*%means

## Effect sizes ##

fInt<-cvInt/sqrt(length(means)*sum(contInt^2*sd^2))

#### expected N
res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=fInt^2)
ceiling(res$v)+length(means)

```

## Guessing the interaction effect size from one-way designs

Consider the case in which the pattern of means in Table 3, case 1, in the paper is taken from a one-way design in which A1 and A2 show means equal to 5 and 2, respectively, and the same within groups variability (say equal to 1). Basically, the researcher observes in the literature a one-way design with only A as a factor and wishes to test the moderating effect of B in a 2 x 2 design. The B factor has two levels, that for simplicity we name “replicated” and “moderated”. The problem is to determine the power parameters of the expected interaction effect.

## Means based method

```{r }
# observed means #
A1<-c(5,2)
# expected means #
A2<-c(0,0)

sd=1
contInt<-c(1,-1,-1,1)

## end input ##

means<-c(A1,A2)
cvInt<-contInt%*%means

## Effect sizes ##

fInt<-cvInt/sqrt(length(means)*sum(contInt^2*sd^2))

#### expected N
res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=fInt^2)
ceiling(res$v)+length(means)


```
### Percentage of  moderation

Researcher should estimate the percentage of moderation, where 100% means full suppression of the effec, 200% full reversing of the effect. 

```{r }
# observed means #
A1<-c(5,2)
cont1<-c(1,-1)
l<-2 #levels of moderators
sd=1
# percentage of moderation
pm<-100

## end of input ##

k1<-length(A1)
k2<-length(A1)*l
f0<-(A1%*%cont1)/(sqrt(length(A1)*sum(cont1^2)*sd^2))

f<-(pm/100)*f0*sqrt(length(cont1)/(length(cont1)*l^2))

#### expected N
(res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=f^2))
ceiling(res$v)+length(cont1)*l

  
```
### Paper example "An example of a more complex design"

Consider a researcher who wishes to design a moderation study based on an one-way design with four conditions implementing an increasing intensity of a stimulus, such that the observed pattern of mean shows a linear trend. In particular, the observed linear trend contrast has an p=.184, corresponding to a f=0.475.


```{r }
### starting from eta-squared ###

# percentage of moderation
pm<-125
# observed parameters #
etap<-.184
cont<-c(-3,-1,1,3)
l<-2 # levels of moderator
### end of input ###

(f0<-sqrt(etap/(1-etap)))

(f<-(pm/100)*f0*sqrt(length(cont1)/(length(cont1)*l^2)))

#### expected N (total sample)
(res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=f^2))
ceiling(res$v)+(length(cont)*l)

```

The same results can be obtained if one specifies anticipating all expected means and pooled standard deviation (tedious method)

```{r }

Replicated<-c(5,9,16,20)
sd<-12.239
Expected<-c(5,3,2,1)
cont<-c(-3,-1,1,3)
## end of input ##
contInt<-c(cont,-cont)
means<-c(Replicated,Expected)

(f<-(contInt%*%means)/sqrt(length(contInt)*sum(contInt^2)*sd^2))

#### expected N
res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=f^2)
ceiling(res$v)+length(cont1)*2

  
```

# Regression Analysis

Any effect in simple and the multiple regression for which the $\eta_p^2$ is available can be evaluated witht he following code. here is an example for a regression with three predictors and a small effect size, according to Cohen's (Cohen, 1988) classification.


```{r }
## input ##
etap<-0.02
k<-3 # number of predictors
## end of input ###
(f2<-etap/(1-etap))
res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=f^2)
ceiling(res$v)+k

```

# Moderated regression 

## Interaction between continuous and dichotomous predictors.

The researcher needs to estimate the expected correlation between the continuous independent variable and the dependent variable for the two groups defined by the moderator.

```{r }
## input ##
ra<-.30
rb<-.50
k<-3 # number of coefficients in the regression (not considering the intercept)
## end of input ##

(bint<-abs(ra-rb))
(f2<-bint^2/(2*(2-ra^2-rb^2)))

## expected N ###
res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=f2)
ceiling(res$v)+k+1


```
## Interaction between continuous predictors

The researcher needs to estimate the expected correlation between the continuous independent variables and the dependent variable, and the change of correlation from the average value of the moderator to one standard deviation above average of the moderator. We assume here that there

```{r }
## input ##
r_yx<-.35
r_ym<-.25
r_change<-.175
k<-3 # number of coefficients in the regression (not considering the intercept)
## end of input ##

(bint<-abs(ra-rb))
(f2<-bint^2/(2*(2-ra^2-rb^2)))

## expected N ###
res<-pwr.f2.test(u=1,power=.80,sig.level = .05, f2=f2)
ceiling(res$v)+k+1


```



# Mediation


### method based on Sobel test
```{r}
if(!require("powerMediation"))
{install.packages("powerMediation")
  library("powerMediation")}
### input ###
# a, b,  are the coefficients as in standard mediational path diagram
a<-.8186
b<-.4039
c<- .4334
### end of input ###

# compute sigma.epsilon
sigma.epsilon<-sqrt(1-(b^2+c^2+2*a*b*c))

## results: expected N
res<-ssMediation.Sobel(power = .80,  theta.1a = .8186, lambda.a = .4039, sigma.x = 1, sigma.m = 1, sigma.epsilon = .6020)
ceiling(res$n)

```

### method based on Sobel test
```{r }
if(!require("bmem"))
  install.packages("bmem")
 library("bmem")
### input ###
# a, b,  are the coefficients as in standard mediational path diagram
a<-.8186
b<-.4039
c<- .4334

### end of input ###

model <- paste(
  'M ~ a*X + start(',a,')*X',
  'Y ~ b*M + c*X + start(',b,') * M + start(.4334)*X',
  'X ~~ start(1)*X',
  'M ~~ start(1)*M',
  'Y ~~ start(1)*Y'
)
set.seed(1234)
power.result <- power.boot(model, indirect = 'ab := a*b', nobs = 100) 
summary(power.result)
## results: expected N
summary
```


