---
output:
  html_document: default
  word_document: default
editor_options: 
  chunk_output_type: console
---

```{r}
library(jtools)

```

# How to do power analysis

## Main features of G*Power

* Cohen's d: $M_1-M_2 \over{SD_{pooled}}$
* t-test to d:  $d={2t \over{sqrt(N-2)}}$

## Paired Groups (repeated measures)

* Cohen's standardized d [bonett2015interval, 367, eq. 1]: $d \over{sqrt{2(1-\rho)}}$
* $\d=.50$, $\rho=.rr$ $d_z$=`r .50/(sqrt(2*(1-.55)))`
* d from t $d_z={t \over{sqrt{N}}}$

## One way Analysis of Variance

From $\eta^2$ to $\epsilon^2$ (Eikeland, 1974, p. 16 eq 9 adapted)
$$\epsilon^2={1-(1-\eta_s^2) \cdot {N-1 \over {N-k}}}$$
Check the numbers
```{r}

eta2<-.35
N<-200
k<-8
(eps<-1-(1-eta2)*((N-1)/(N-k)))

```







## Factorial Designs

$$\eta_p^2={{\sigma_f^2} \over{\sigma_f^2+\sigma^2}}$$

Eikeland, 1974, p. 16 eq 9 adapted: let $df$ be the degrees of freedom of the effect of interest, and $K$ the **total** number of groups in the design and $N$ the total sample, 

$$\epsilon_p^2={1-(1-\eta_p^2) \cdot {N-K+df \over {N-K}}}$$

Example from Eikeland, 1974, p. 16 
```{r}
N<-20
k<-4
df<-1
N-k+df
eta2<-.20
1-(1-eta2)*((N-k+df)/(N-k))


```

$$\epsilon_p^2={1-(1-.20) \cdot {20-4+1 \over {20-4}}}=1-.80 \cdot {17 \over 16}=.15$$


Porportions of variances: $\delta_e=.40$ $\delta_r=1-.10-.40=.50$ \eta_p^2$=`r .10/.60`.

A check epsilon-squared

```{r}

a<-rep(c(0,1),20)
b<-rep(c(0,1),each=20)
y<-a+a*b+rnorm(40)
a<-factor(a)
b<-factor(b)
model<-lm(y~a*b)
library(car)
ss<-Anova(model, type = 3)
ssef<-ss$`Sum Sq`[2]
sserr<-ss$`Sum Sq`[5]
dferr<-ss$Df[5]
mse<-sserr/dferr
ka<-length(levels(a))
(eta2<-(ssef/(ssef+sserr)))
(eps1<-(ssef-(ka-1)*mse)/(ssef+sserr))
N<-length(a)
K<-length(levels(a))*length(levels(b))
(eps2<-1-(1-eta2)*(N-K+ka-1)/(N-K))

```

## Power analysis for contrasts

Adapted from Gpower Manual 3.0, pag 29. 
$$f={ |\sum{ c_i \cdot u_i}| \over{ \sqrt{k \cdot \sum{c_i^2 \cdot \sigma^2}}}}$$

check the numbers
```{r}

10/sqrt(4*4*5^2)

```
## Guessing the interaction effect size from one-way designs

$$f_n={{p_m \over 100} \cdot f_o \cdot \sqrt{{k_0 \over{k_n \cdot l} }}}$$


$$f_n={{p_m \over 100} \cdot f_o \cdot {1 \over 2}}$$

Examples: `r 1.50*1/2` `r 2*3/2*1/2`,`r .50*.50*1.5`




```{r}
# table 3 data case1
meansA<-c(5,2)
meansAB<-c(5,2,0,0)
contA<-c(1,-1)
(val<-abs(meansA%*%contA))
(f<-val/(sqrt(2*sum(contA^2))))

contAB<-c(1,-1,-1,1)
(val<-abs(meansAB%*%contAB))
(f<-val/(sqrt(4*sum(contAB^2))))

# table 3 data case2
meansAB<-c(5,2,2,5)
contAB<-c(1,-1,-1,1)
(val<-abs(meansAB%*%contAB))
(f<-val/(sqrt(4*sum(contAB^2))))

```

# An example of more complex designs 



```{r}

es<-.475
pm<-1.25

cont<-c(-3,-1,1,3)
m<-c(5,9,16,20)
md<-as.data.frame(m)
md$g<-c("s1","s2","s3","s4")
library(ggplot2)
gp<-ggplot(md,aes(y=m,x=g,group=1))
gp + geom_point()+geom_line()
sigma<-12.239
(cc<-cont%*%m)
(f<-cc/sqrt(length(cont)*sum(cont^2)*sigma^2))


cont2<-c(-3,-1,1,3,3,1,-1,-3)
m2<-c(5,9,16,20,5,3,2,1)
md2<-as.data.frame(m2)
md2$g<-c("s1","s2","s3","s4")
md2$g2<-c(rep("Replicated",4),rep("Moderated",4))
md2
(cc2<-cont2%*%m2)
(f2<-cc2/sqrt(length(cont2)*sum(cont2^2*sigma^2)))
gp<-ggplot(md2,aes(y=m2,x=g,group=g2,shape=g2))
gp<-gp + geom_point(size=5)+geom_line()
gp<-gp+xlab("Stimulus intensity")+ylab("Dependent variable")+labs(shape="Moderator")
gp<-gp + expand_limits(y = c(-5,25))
gp
(esn<-es*pm*(sqrt(length(cont)/(length(cont2)*2))))



## check
contx<-c(-3,-1,1,3)
mx<-c(5,3,2,1)
(ccx<-contx%*%mx)
(fx<-ccx/sqrt(length(contx)*sum(contx^2)*sigma^2))
(abs(fx)+f)/f



```

$$f_n={1.25 \cdot .475 \cdot {\sqrt{4 \over {8 \cdot 2}}}}=0.296$$



## Interaction between continuous and dichotomous predictors

```{r}
ra<-.50
rb<-.00
(r<-mean(ra,rb))

ri<-ra-rb
(sigma<-sqrt((2-(ra^2+rb^2))/2))

(ff<-ri^2/(2*(2-ra^2-rb^2)))


```

$$f^2 \approx {\beta_{int}^2 \over{{2 \cdot (2-r_a^2-r_b^2)}}}$$

## Interaction between continuous

```{r}

### example in the paper
library(pwr)
 rx<-.35
 rm<-.25
 pm<-.50
 (betai<-rx*pm)
# 
 (err<-1-rx^2-rm^2)

 
 (ff<-betai^2/err)
 library(pwr)
 (pp<-pwr.f2.test(u=1,f2=ff,power=.80))
 #required N
 round(pp$v+3)
# 
# #####


```

$$f^2 \approx {\beta_{int}^2 \over{{1 - r_{yx}^2-r_{ym}^2}}}$$

## Mediation analysis









